---
title: "Class 11: Structural Bioinformatics (Cont.)"
author: "Hailey Wheeler (A13312713)"
format: html
---

Today we are goign to finish off Lab 10 on learning about and analyzing protein structures

## Section 4: Comparative structure analysis of Adenylate Kinase
get.seq(): Input-database identifier 
blast.pdb()
get.pdb()
pdbaln(): Alignment of all structures 
pca(), plot()

Use 'bio3d' for analysis that starts with a single sequence
'msa' package from BioC
- Install 'BiocManager' and use 'BiocManager::install()'

```{r}
library(bio3d)
```

First, pick sequence of interest '1ake_A'
```{r}
aa <- get.seq("1ake_A")
aa
```

search for all related structures in the PDB database 

```{r}
b <- blast.pdb(aa)
```

```{r}
plot(b)
```

```{r}
hits <- plot(b)
```
```{r}
hits$pdb.id
```

Lets save our results thus far so we dont have to run blast again. 

```{r}
#blast.pdb("1AKE_1")

save(hits, b, file="blast_results.Rds")
load("blast_results.Rds")
hits
```

Download all these related structures from the database with 'get.pdb()'
```{r}
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```
```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

```{r}
pdbs
```

## Principal Component Analysis 
We can do a PCA on the coordinate data (i.e. the '$xyz' values) of all the structures

```{r}
pc.xray <- pca(pdbs)
plot(pc.xray)
```

```{r}
dim(pdbs$xyz)
```

```{r}
# Calculate RMSD
rd <- rmsd(pdbs)

# Structure-based clustering
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)

plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```


```{r}
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```


```{r}
#Plotting results with ggplot2
library(ggplot2)
library(ggrepel)

ids <- basename.pdb(pdbs$id)

df <- data.frame(PC1=pc.xray$z[,1], 
                 PC2=pc.xray$z[,2], 
                 col=as.factor(grps.rd),
                 ids=ids)

p <- ggplot(df) + 
  aes(PC1, PC2, col=col, label=ids) +
  geom_point(size=2) +
  geom_text_repel(max.overlaps = 20) +
  theme(legend.position = "none")
p
```

```{r}
modes <- nma(pdbs)
plot(modes, pdbs, col=grps.rd)
```









