---
title: "Class 08: Halloween Mini-Project"
author: "Hailey Wheeler (A13312713)"
format: pdf
---

```{r}
candy_file <- "candy-data.csv"
candy = read.csv(candy_file, row.names = 1)
head(candy)
```

Question 1: How many different candy types are in this dataset? 
```{r}
nrow(candy)
```

Question 2: How many fruity candy types are in the dataset? 
```{r}
sum(candy$fruity)
```

Question 3: What is your favorite candy in the dataset and what is it's winpercent value? 

```{r}
candy["Almond Joy", ]$winpercent
```

Question 4: What is the winpercent value for "Kit Kat"?
```{r}
candy["Kit Kat",]$winpercent
```

Question 5: What is the winpercent value for "Tootsie Roll Snack Bars"? 
```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

```{r}
library(skimr)
skim(candy)
```

Question 6: Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, winpercent is on a different scale compared to the other columns. 

Question 7: What do you think a zero and one represent for the candy$chocolate column?

The zero and one indicate if it is a chocolate-y candy (=1), or not (=0).

Question 8: Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy, aes(x=winpercent)) +
  geom_histogram()
```


Question 9: Is the distribution of winpercent values symmetrical?

No. 

Question 10: Is the center of the distribution above or below 50%?

Below. 

Question 11: On average is chocolate candy higher or lower ranked than fruit candy? 

Higher. 

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])

```

Question 12:. Is this difference statistically significant?

Yes. 

```{r}
choc <- candy$winpercent[as.logical(candy$chocolate)]
fruit <- candy$winpercent[as.logical(candy$fruity)]
t.test(choc, fruit)
```


Question 13: What are the five least liked candy types in this set?

Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters. 

Question 14: What are the top 5 all time favorite candy types out of this set?

Reese's Peanut Butter Cup, Reese's Miniatures, Twix, Kit Kat, and Snickers.

 
```{r}
head(candy[order(candy$winpercent, decreasing = TRUE), ], n=5)
```

```{r}
head(candy[order(candy$winpercent), ], n=5)
```

Question 15: Make a first barplot of candy ranking based on winpercent values. 

```{r}
ggplot(candy) + 
  aes(x=winpercent, y=rownames(candy)) +
  geom_col()
```

Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) + 
  geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)]= "brown"
my_cols[as.logical(candy$fruity)] = "pink"
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) + 
  geom_col(fill=my_cols)
```

Question 17: What is the worst ranked chocolate candy?

Boston Baked Beans.

Question 18: What is the best ranked fruity candy?

Starbursts.

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) #+ 
  #geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

Question 19: Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Tootsie Roll Midgies.

Question 20: What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nip (least popular), Nestle Smarties, Ring Pop, Mr. Good Bar, Hershey's Milk Chocolate

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Question 21: Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().
```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```


```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```
Question 22: Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity.

Question 23:  Similarly, what two variables are most positively correlated?

Winpercent and chocolate. 

```{r}
cor(candy$chocolate, candy$bar)
cor(candy$chocolate, candy$winpercent)
```

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + #geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
#library(plotly)
#ggplotly(p)
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

Question 24: What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? 

Pluribus and Fruity.
