---
title: "Class 14: Pathway Analysis from RNA-Seq"
author: "Hailey Wheeler (A13312713)"
format: pdf
toc: true
---


```{r}
library(DESeq2)
```

```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"
```

```{r}
colData = read.csv(metaFile, row.names=1)
head(colData)
```

## I want to get rid of the first "length" column in counts 
```{r}
countData = read.csv(countFile, row.names=1)
head(countData)
colnames(countData)
```
Q. Complete the code below to remove the troublesome first column from countData

```{r}
countData <- as.matrix(countData[,-1])
head(countData)
colnames(countData)
dim(countData)
```
## Remove zeros
Q. Complete the code below to filter countData to exclude genes (i.e. rows) where we have 0 read count across all samples (i.e. columns).
```{r}
to.rm.ind <- rowSums(countData) == 0
countData <- countData[!to.rm.ind,]
nrow(countData)
```

## Run DESeq2

```{r}
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)
```

```{r}
dds
```

```{r}
res = results(dds)
```

Q. Call the summary() function on your results to get a sense of how many genes are up or down-regulated at the default 0.1 p-value cutoff.
```{r}
summary(res)
```

```{r}
plot( res$log2FoldChange, -log(res$padj) )
```

## Improve Plot below 
Q. Improve this plot by completing the below code, which adds color and axis labels

```{r}
# Make a color vector for all genes
mycols <- rep("gray", nrow(res) )

# Color red the genes with absolute fold change above 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Color blue those with adjusted p-value less than 0.01
#  and absolute fold change more than 2
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col= mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
columns(org.Hs.eg.db)
```

Q. Use the mapIDs() function multiple times to add SYMBOL, ENTREZID and GENENAME annotation to our results by completing the code below
```{r}
##res$symbol = mapIds(org.Hs.eg.db,
                   # keys=row.names(counts), 
                   # keytype="ENSEMBL",
                   # column="SYMBOL",
                   # multiVals="first")

##res$entrez = mapIds(org.Hs.eg.db,
                   # keys=row.names(counts),
                   # keytype="ENSEMBL",
                   # column="ENTREZID",
                   # multiVals="first")

#res$name =   mapIds(org.Hs.eg.db,
                  #  keys=row.names(counts),
                  #  keytype="ENSEMBL",
                 #   column="GENENAME",
                   # multiVals="first")

# head(res, 10)
```

Q. Finally for this section let's reorder these results by adjusted p-value and save them to a CSV file in your current project directory.
```{r}
res = res[order(res$pvalue),]
write.csv(res, file ="deseq_results.csv")
```

```{r}
library(pathview)
```
```{r}
library(gage)
library(gageData)
```

```{r}
data(kegg.sets.hs)
data(sigmet.idx.hs)
```

```{r}
# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]
```

```{r}
# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
attributes(keggres)
```

```{r}
head(keggres$less)
```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```


```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
```


```{r}
## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```


```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

Q. Can you do the same procedure as above to plot the pathview figures for the top 5 down-regulated pathways? Yes. 

```{r}
downkeggrespathways <- rownames(keggres$less)[1:5]
downkeggresids = substr(downkeggrespathways, start=1, stop=8)
pathview(gene.data=foldchanges, pathway.id=downkeggresids, species="hsa")
```


## Gene Ontology 

```{r}
data(go.sets.hs)
data(go.subs.hs)
```

```{r}
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```

```{r}
#sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
#print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
#write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

Q: What pathway has the most significant “Entities p-value”? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?

The pathway with the most significant "Entities p-value" is 'Cell Cycle'. This is identical to the KEGG results. The main factor that can cause differences between the two methods is that Reactome has a larger profile and increased specificity. 




